---
title: "Creating Task Objects in mcdaHub"
author: "Wanting Zuo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Task Objects in mcdaHub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, include = FALSE}
library(mcdaHub)
# Load datasets for vignette
# Use data() in a new environment to avoid conflicts with R's built-in datasets
env <- new.env()
data(cars, package = "mcdaHub", envir = env)
data(supplier, package = "mcdaHub", envir = env)
data(employee, package = "mcdaHub", envir = env)
cars_data <- env$cars
supplier_data <- env$supplier
employee_data <- env$employee
```

# Creating Task Objects in mcdaHub

This vignette demonstrates how to create `TaskRanking`, `TaskSorting`, and `TaskChoice` objects in the mcdaHub package.

## Overview

Task objects define your decision problem. There are three types:

- **TaskRanking**: Rank alternatives from best to worst
- **TaskChoice**: Select the best alternative(s)
- **TaskSorting**: Assign alternatives to predefined categories

## TaskRanking Class

`TaskRanking` is used to rank alternatives from best to worst.

### Creating TaskRanking Object

```{r task-ranking-create}
# Create a ranking task
# (cars_data is already loaded in the setup chunk)
task_ranking <- TaskRanking$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Display the task
print(task_ranking)
```

### Parameters

- `alt`: Character vector of alternative names
- `crit`: Character vector of criterion names
- `perf`: Numeric matrix (alternatives × criteria) - performance values
- `weight`: Numeric vector of criterion weights (should sum to 1)
- `direction`: Character vector ("min" or "max" for each criterion)

### Example: Viewing Task Components

```{r task-ranking-components}
# Access task components
cat("Alternatives:\n")
print(task_ranking$alt)

cat("\nCriteria:\n")
print(task_ranking$crit)

cat("\nPerformance Matrix:\n")
print(task_ranking$perf)

cat("\nWeights:\n")
print(task_ranking$weight)

cat("\nDirections:\n")
print(task_ranking$direction)
```

## TaskChoice Class

`TaskChoice` is used to select the best alternative(s).

### Creating TaskChoice Object

```{r task-choice-create}
# Create a choice task (same structure as ranking)
task_choice <- TaskChoice$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Display the task
print(task_choice)
```

### Parameters

Same as `TaskRanking`:
- `alt`: Character vector of alternative names
- `crit`: Character vector of criterion names
- `perf`: Numeric matrix (alternatives × criteria)
- `weight`: Numeric vector of criterion weights
- `direction`: Character vector ("min" or "max" for each criterion)

**Note:** `TaskChoice` uses the same input format as `TaskRanking`, but the result interpretation differs (selects best alternative vs. full ranking).

## TaskSorting Class

`TaskSorting` is used to assign alternatives to predefined categories.

### Method 1: Traditional Sorting (with Profile Matrix)

```{r task-sorting-traditional}
# Create a sorting task with profile matrix
# Ensure prof is a numeric matrix (not a vector or data.frame)
prof_matrix <- as.matrix(cars_data$prof)
# Ensure it's numeric (as.matrix() might create character matrix if data has factors)
if (!is.numeric(prof_matrix)) {
  prof_matrix <- matrix(as.numeric(prof_matrix), nrow = nrow(prof_matrix), ncol = ncol(prof_matrix))
  rownames(prof_matrix) <- rownames(cars_data$prof)
  colnames(prof_matrix) <- colnames(cars_data$prof)
}
task_sorting <- TaskSorting$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction,
  cat_names = cars_data$cat_names,
  prof = prof_matrix,
  assign_rule = "pessimistic"
)

# Display the task
print(task_sorting)
```

### Parameters for Traditional Sorting

- `alt`: Character vector of alternative names
- `crit`: Character vector of criterion names
- `perf`: Numeric matrix (alternatives × criteria)
- `weight`: Numeric vector of criterion weights
- `direction`: Character vector ("min" or "max" for each criterion)
- `cat_names`: Character vector of category names (e.g., c("Low", "Medium", "High"))
- `prof`: Numeric matrix (profiles × criteria) - profile performance values
- `assign_rule`: "pessimistic" or "optimistic" assignment rule

### Method 2: AHPSort (with Pairwise Comparisons)

```{r task-sorting-ahp}
# Create a sorting task with AHPSort data
# Note: For AHPSort, prof is not needed - explicitly set to NULL
# prof_order is optional metadata, not required for TaskSorting initialization
# Only pass the fields that employee_data contains (no prof field)
task_sorting_ahp <- TaskSorting$new(
  alt = employee_data$alt,
  crit = employee_data$crit,
  cat_names = employee_data$cat_names,
  pair_crit = employee_data$pair_crit,
  judge_alt_profile = employee_data$judge_alt_profile,
  prof = NULL,  # Explicitly set to NULL for AHPSort
  assign_rule = "pessimistic"
)

# Display the task
print(task_sorting_ahp)
```

### Parameters for AHPSort

- `alt`: Character vector of alternative names
- `crit`: Character vector of criterion names
- `cat_names`: Character vector of category names
- `pair_crit`: Pairwise comparison matrix for criteria (must have rownames and colnames)
- `judge_alt_profile`: Data.frame with columns: criterion, alt, profile, value
- `prof_order`: Character vector of profile order (from worst to best)
- `assign_rule`: "pessimistic" or "optimistic" assignment rule

### Example: Viewing AHPSort Task Components

```{r task-sorting-ahp-components}
# Access task components
cat("Alternatives:\n")
print(task_sorting_ahp$alt)

cat("\nCriteria:\n")
print(task_sorting_ahp$crit)

cat("\nCategory Names:\n")
print(task_sorting_ahp$cat_names)

cat("\nProfile Order (from employee_data, not stored in TaskSorting):\n")
print(employee_data$prof_order)

cat("\nCriteria Pairwise Comparison Matrix:\n")
print(task_sorting_ahp$pair_crit)

cat("\nAlternative-Profile Judgments:\n")
print(head(task_sorting_ahp$judge_alt_profile))
```

## Summary

| Task Type | Purpose | Key Parameters |
|-----------|---------|----------------|
| `TaskRanking` | Rank alternatives | `alt`, `crit`, `perf`, `weight`, `direction` |
| `TaskChoice` | Select best alternative(s) | Same as `TaskRanking` |
| `TaskSorting` | Assign to categories | Traditional: `prof`, `cat_names`<br>AHPSort: `pair_crit`, `judge_alt_profile`, `prof_order` |

## Next Steps

After creating a task object, you can:
1. Use a `Decider` object to solve it (see `vignette("02-decider-and-solve", package = "mcdaHub")`)
2. Validate the task using `task$validate()`
3. Access task components for inspection

