---
title: "Creating Decider Objects and Using solve() in mcdaHub"
author: "Wanting Zuo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Decider Objects and Using solve() in mcdaHub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, include = FALSE}
library(mcdaHub)
# Load datasets for vignette
# Use data() in a new environment to avoid conflicts with R's built-in datasets
env <- new.env()
data(cars, package = "mcdaHub", envir = env)
data(supplier, package = "mcdaHub", envir = env)
data(employee, package = "mcdaHub", envir = env)
cars_data <- env$cars
supplier_data <- env$supplier
employee_data <- env$employee
```

# Creating Decider Objects and Using solve() in mcdaHub

This vignette demonstrates how to create `DeciderTOPSIS`, `DeciderPROMETHEE`, and `DeciderAHP` objects and use their `solve()` method to solve MCDA tasks.

## Overview

Decider objects implement MCDA algorithms. There are three types:

- **DeciderTOPSIS**: TOPSIS method (Technique for Order Preference by Similarity to Ideal Solution)
- **DeciderPROMETHEE**: PROMETHEE II method (Preference Ranking Organization Method)
- **DeciderAHP**: AHP method (Analytic Hierarchy Process) and AHPSort

## DeciderTOPSIS Class

### Creating DeciderTOPSIS Object

```{r decider-topsis-create}
# Create a TOPSIS decider
decider_topsis <- DeciderTOPSIS$new()

# The decider is ready to use
class(decider_topsis)
```

### Using solve() for Ranking Tasks

```{r decider-topsis-ranking}
# Create a ranking task
task_ranking <- TaskRanking$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Create decider
decider_topsis <- DeciderTOPSIS$new()

# Solve the task
result <- decider_topsis$solve(task_ranking)

# View results
print(result)

# Access ranking table
result$ranking_table
```

### Using solve() for Choice Tasks

```{r decider-topsis-choice}
# Create a choice task
task_choice <- TaskChoice$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Create decider
decider_topsis <- DeciderTOPSIS$new()

# Solve the choice task
result_choice <- decider_topsis$solve(task_choice)

# View results
print(result_choice)

# Access selected alternatives
result_choice$choice_table
```

### Using solve() for Sorting Tasks

```{r decider-topsis-sorting}
# Create a sorting task
# Ensure prof is a numeric matrix (not a vector or data.frame)
prof_matrix <- as.matrix(cars_data$prof)
# Ensure it's numeric (as.matrix() might create character matrix if data has factors)
if (!is.numeric(prof_matrix)) {
  prof_matrix <- matrix(as.numeric(prof_matrix), nrow = nrow(prof_matrix), ncol = ncol(prof_matrix))
  rownames(prof_matrix) <- rownames(cars_data$prof)
  colnames(prof_matrix) <- colnames(cars_data$prof)
}
task_sorting <- TaskSorting$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction,
  cat_names = cars_data$cat_names,
  prof = prof_matrix,
  assign_rule = "pessimistic"
)

# Create decider
decider_topsis <- DeciderTOPSIS$new()

# Solve the sorting task
result_sorting <- decider_topsis$solve(task_sorting)

# View results
print(result_sorting)

# Access sorting table
result_sorting$sorting_table
```

## DeciderPROMETHEE Class

### Creating DeciderPROMETHEE Object

```{r decider-promethee-create}
# Create a PROMETHEE decider
decider_promethee <- DeciderPROMETHEE$new()

# The decider is ready to use
class(decider_promethee)
```

### Using solve() for Ranking Tasks

```{r decider-promethee-ranking}
# Create a ranking task
task_ranking <- TaskRanking$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Create decider
decider_promethee <- DeciderPROMETHEE$new()

# Solve the task
result <- decider_promethee$solve(task_ranking)

# View results
print(result)

# Access metadata (leaving flow, entering flow, net flow)
result$metadata
```

### Using solve() for Choice Tasks

```{r decider-promethee-choice}
# Create a choice task
task_choice <- TaskChoice$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Create decider
decider_promethee <- DeciderPROMETHEE$new()

# Solve the choice task
result_choice <- decider_promethee$solve(task_choice)

# View results
print(result_choice)
```

### Using solve() for Sorting Tasks

```{r decider-promethee-sorting, eval = FALSE}
# Create a sorting task
# Ensure prof is a numeric matrix (not a vector or data.frame)
prof_matrix <- as.matrix(cars_data$prof)
# Ensure it's numeric (as.matrix() might create character matrix if data has factors)
if (!is.numeric(prof_matrix)) {
  prof_matrix <- matrix(as.numeric(prof_matrix), nrow = nrow(prof_matrix), ncol = ncol(prof_matrix))
  rownames(prof_matrix) <- rownames(cars_data$prof)
  colnames(prof_matrix) <- colnames(cars_data$prof)
}
task_sorting <- TaskSorting$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction,
  cat_names = cars_data$cat_names,
  prof = prof_matrix
)

# Create decider
decider_topsis <- DeciderTOPSIS$new()

# Solve the sorting task 
result_sorting <- decider_topsis$solve(task_sorting)

# View results
print(result_sorting)

# Access sorting table
result_sorting$sorting_table
```

## DeciderAHP Class

### Creating DeciderAHP Object

```{r decider-ahp-create}
# Create an AHP decider
decider_ahp <- DeciderAHP$new()

# The decider is ready to use
class(decider_ahp)
```

### Using solve() for Ranking Tasks (AHP)

```{r decider-ahp-ranking}
# Create a ranking task with pairwise comparisons
task_ranking_ahp <- TaskRanking$new(
  alt = supplier_data$alt,
  crit = supplier_data$crit,
  pair_crit = supplier_data$pair_crit,
  pair_alt = supplier_data$pair_alt
)

# Create decider
decider_ahp <- DeciderAHP$new()

# Solve with AHP
result_ranking_ahp <- decider_ahp$solve(task_ranking_ahp)

# View results
print(result_ranking_ahp)

# Access metadata (criteria weights, consistency ratio)
result_ranking_ahp$metadata
```

### Using solve() for Choice Tasks (AHP)

```{r decider-ahp-choice}
# Create a choice task with pairwise comparisons
task_choice_ahp <- TaskChoice$new(
  alt = supplier_data$alt,
  crit = supplier_data$crit,
  pair_crit = supplier_data$pair_crit,
  pair_alt = supplier_data$pair_alt
)

# Create decider
decider_ahp <- DeciderAHP$new()

# Solve with AHP
result_choice_ahp <- decider_ahp$solve(task_choice_ahp)

# View results
print(result_choice_ahp)
```

### Using solve() for Sorting Tasks (AHPSort)

```{r decider-ahp-sorting}
# Create a sorting task with AHPSort data
# Note: For AHPSort, prof is not needed - explicitly set to NULL
# prof_order is optional metadata, not required for TaskSorting initialization
# Only pass the fields that employee_data contains (no prof field)
task_sorting_ahp <- TaskSorting$new(
  alt = employee_data$alt,
  crit = employee_data$crit,
  cat_names = employee_data$cat_names,
  pair_crit = employee_data$pair_crit,
  judge_alt_profile = employee_data$judge_alt_profile,
  prof = NULL,  # Explicitly set to NULL for AHPSort
  assign_rule = "pessimistic"
)

# Create decider
decider_ahp <- DeciderAHP$new()

# Solve with AHP (AHPSort method)
result_sorting_ahp <- decider_ahp$solve(task_sorting_ahp)

# View results
print(result_sorting_ahp)

# Access sorting table
result_sorting_ahp$sorting_table
```

## Understanding Results

### ResultRanking

```{r understanding-result-ranking}
# Solve a ranking task
task <- TaskRanking$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

result <- DeciderTOPSIS$new()$solve(task)

# Access ranking table
ranking_table <- result$ranking_table
print(ranking_table)

# Access metadata
metadata <- result$metadata
names(metadata)
```

### ResultChoice

```{r understanding-result-choice}
# Solve a choice task
task_choice <- TaskChoice$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

result_choice <- DeciderTOPSIS$new()$solve(task_choice)

# Access choice table
choice_table <- result_choice$choice_table
print(choice_table)
```

### ResultSorting

```{r understanding-result-sorting}
# Solve a sorting task
task_sorting <- TaskSorting$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction,
  cat_names = cars_data$cat_names,
  prof = {
    pm <- as.matrix(cars_data$prof)
    if (!is.numeric(pm)) {
      pm <- matrix(as.numeric(pm), nrow = nrow(pm), ncol = ncol(pm))
      rownames(pm) <- rownames(cars_data$prof)
      colnames(pm) <- colnames(cars_data$prof)
    }
    pm
  },
  assign_rule = "pessimistic"
)

result_sorting <- DeciderTOPSIS$new()$solve(task_sorting)

# Access sorting table
sorting_table <- result_sorting$sorting_table
print(sorting_table)
```

## Summary

| Decider | Ranking | Choice | Sorting | Notes |
|---------|---------|--------|---------|-------|
| `DeciderTOPSIS` | ✓ | ✓ | ✓ | Uses performance matrix |
| `DeciderPROMETHEE` | ✓ | ✓ | ✓ | Uses performance matrix, FlowSort for sorting |
| `DeciderAHP` | ✓ | ✓ | ✓ | Uses pairwise comparisons, AHPSort for sorting |

## Common Workflow

1. **Create a Task object** (see `vignette("01-task-classes", package = "mcdaHub")`)
2. **Create a Decider object**: `decider <- DeciderTOPSIS$new()`
3. **Solve the task**: `result <- decider$solve(task)`
4. **View results**: `print(result)`
5. **Access result components**: `result$ranking_table`, `result$metadata`, etc.

## Next Steps

- Learn about sensitivity analysis: `vignette("03-analysis-tools", package = "mcdaHub")`
- Compare different algorithms using `AlgoRankCorrelation`

