---
title: "Analysis Tools: WeightPerturbation and AlgoRankCorrelation in mcdaHub"
author: "Wanting Zuo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysis Tools: WeightPerturbation and AlgoRankCorrelation in mcdaHub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

```{r setup, include = FALSE}
library(mcdaHub)
library(ggplot2)
# Load datasets for vignette
# Use data() in a new environment to avoid conflicts with R's built-in datasets
env <- new.env()
data(cars, package = "mcdaHub", envir = env)
cars_data <- env$cars
```

# Analysis Tools: WeightPerturbation and AlgoRankCorrelation in mcdaHub

This vignette demonstrates how to use `WeightPerturbation` for sensitivity analysis and `AlgoRankCorrelation` for comparing different MCDA algorithms.

## Part 1: WeightPerturbation Class

`WeightPerturbation` performs sensitivity analysis through weight perturbation to understand how robust your MCDA results are to changes in criterion weights.

### Overview

The `WeightPerturbation` class implements One-At-A-Time (OAT) weight perturbation:
- Varies one criterion weight at a time
- Keeps other weights proportional
- Analyzes how rankings change across perturbation scenarios

### Creating WeightPerturbation Object

```{r weight-perturbation-create}
# Create a ranking task
task <- TaskRanking$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Create a decider
decider <- DeciderTOPSIS$new()

# Create WeightPerturbation object
wp <- WeightPerturbation$new(
  task = task,
  decider = decider,
  perturb_rg = 0.3,  
  perturb_n = 50     
)

# Display the object
print(wp)
```

**Key Parameters:**
- `task`: TaskRanking object (only ranking tasks are supported)
- `decider`: Decider object (DeciderTOPSIS, DeciderPROMETHEE, etc.)
- `perturb_rg`: Perturbation range (0.1 to 0.5) - how much to vary weights
- `perturb_n`: Number of steps per criterion (50 to 200) - resolution of analysis

### Using weight_perturb() Method

```{r weight-perturbation-run}
# Perform weight perturbation analysis
wp$weight_perturb()

# This method:
# 1. Solves the task with original weights (baseline)
# 2. Generates perturbed weight sets using OAT method
# 3. Solves the task for each perturbed weight set
# 4. Stores all results for further analysis
```

### Using perturb_rank() Method

```{r perturb-rank}
# Extract rank matrix (scenarios × alternatives)
rank_mat <- wp$perturb_rank()

# View the rank matrix
head(rank_mat)

# Dimensions: (scenarios × alternatives)
dim(rank_mat)

# Each row represents a perturbation scenario
# Each column represents an alternative
# Values are ranks (1 = best, higher numbers = worse)
```

**What perturb_rank() returns:**
- Matrix with dimensions (scenarios × alternatives)
- Each row: one perturbation scenario
- Each column: one alternative
- Values: ranks (1 = best)

### Using perturb_stab_plot() Method

```{r perturb-stab-plot-trajectory, eval = FALSE}
# Rank trajectory plot
# Shows how each alternative's rank changes across scenarios
p1 <- wp$perturb_stab_plot(type = "trajectory")
print(p1)
```

```{r perturb-stab-plot-sd, eval = FALSE}
# Standard deviation plot
# Shows standard deviation of ranks for each alternative
p2 <- wp$perturb_stab_plot(type = "sd")
print(p2)
```

**Plot Types:**
- `type = "trajectory"`: Shows how each alternative's rank changes across scenarios (line plot)
- `type = "sd"`: Shows standard deviation of ranks for each alternative (bar plot)

**Interpretation:**
- **Trajectory plot**: Flat lines indicate stable rankings; wavy lines indicate sensitivity
- **SD plot**: Lower SD values indicate more stable rankings

### Using perturb_corr_heatmap() Method

```{r perturb-corr-heatmap, eval = FALSE}
# Correlation heatmap between perturbation scenarios
# Shows how similar rankings are across different weight perturbations
heatmap <- wp$perturb_corr_heatmap(method = "spearman")
print(heatmap)
```

**Parameters:**
- `method`: Correlation method ("spearman", "kendall", or "pearson")

**What it shows:**
- Heatmap of correlations between different perturbation scenarios
- High correlation (red) = similar rankings
- Low correlation (blue) = different rankings

### Using perturb_rank_corr() Method

```{r perturb-rank-corr, eval = FALSE}
# Calculate rank correlation matrix between scenarios
# Note: This requires weight_perturb() to be run first
corr_matrix <- wp$perturb_rank_corr(method = "spearman")

# View correlation matrix
head(corr_matrix)

# Dimensions: (scenarios × scenarios)
dim(corr_matrix)
```

**What it returns:**
- Correlation matrix (scenarios × scenarios)
- Shows how similar rankings are between different perturbation scenarios

## Part 2: AlgoRankCorrelation Class

`AlgoRankCorrelation` compares rankings from different MCDA algorithms to assess consistency.

### Overview

The `AlgoRankCorrelation` class:
- Takes a TaskRanking and a list of Decider objects
- Solves the task with each decider
- Computes pairwise Spearman correlations between rankings
- Helps assess whether different methods produce consistent results

### Creating AlgoRankCorrelation Object

```{r algo-correlation-create}
# Create a ranking task
task <- TaskRanking$new(
  alt = cars_data$alt,
  crit = cars_data$crit,
  perf = cars_data$perf,
  weight = cars_data$weight,
  direction = cars_data$direction
)

# Create multiple deciders
decider1 <- DeciderTOPSIS$new()
decider2 <- DeciderPROMETHEE$new()

# Create AlgoRankCorrelation object
analyzer <- AlgoRankCorrelation$new(
  task = task,
  decider = list(decider1, decider2)
)
```

**Key Parameters:**
- `task`: TaskRanking object (only ranking tasks are supported)
- `decider`: List of Decider objects (at least 2 required)

### Using calculate() Method

```{r algo-correlation-calculate}
# Calculate correlation matrix between algorithm rankings
corr_matrix <- analyzer$calculate()

# View correlation matrix
print(analyzer)

# Access the correlation matrix directly
analyzer$correlation_matrix
```

**What calculate() does:**
1. Solves the task with each decider to get ResultRanking objects
2. Extracts rankings from all ResultRanking objects
3. Validates that all results contain the same alternatives
4. Computes pairwise Spearman correlations between rankings
5. Returns correlation matrix (algorithms × algorithms)

**Correlation values:**
- **1.0**: Perfect positive correlation (identical rankings)
- **0.0**: No correlation (random relationship)
- **-1.0**: Perfect negative correlation (opposite rankings)
- **> 0.7**: Strong positive correlation (consistent results)
- **< 0.3**: Weak correlation (inconsistent results)

### Using print() Method

```{r algo-correlation-print}
# Print correlation matrix
print(analyzer)
```

## Understanding Results

### Weight Perturbation Results

**High rank stability means:**
- Rankings don't change much with weight variations
- Your decision is robust to weight uncertainty
- Weights are well-calibrated

**Low rank stability means:**
- Rankings are sensitive to weight changes
- You should carefully justify your weights
- Consider collecting more information or using weight elicitation methods

### Algorithm Correlation Results

**High correlation (close to 1) means:**
- Algorithms produce similar rankings
- Results are consistent across methods
- Increased confidence in the decision

**Low correlation means:**
- Algorithms produce different rankings
- Results depend on the method chosen
- Consider the problem characteristics when choosing a method
- May indicate that the problem is ambiguous or weights are not well-defined

## Best Practices

### Weight Perturbation

1. **Choose Appropriate Range**: 
   - `perturb_rg = 0.1-0.3`: Small variations (recommended for most cases)
   - `perturb_rg = 0.3-0.5`: Large variations (stress testing)

2. **Sufficient Steps**: 
   - `perturb_n = 51`: Good balance between resolution and computation time
   - `perturb_n = 101+`: Smoother curves, slower computation

3. **Interpret with Context**: 
   - High sensitivity doesn't always mean bad weights
   - Consider your problem's characteristics
   - Use domain knowledge to validate results

4. **Compare Methods**: 
   - Different methods may have different sensitivities
   - Consistent results across methods increase confidence

### Algorithm Comparison

1. **Use Multiple Methods**: Compare at least 2-3 different algorithms

2. **Interpret Correlations**: 
   - High correlation increases confidence
   - Low correlation suggests method-dependent results

3. **Consider Problem Type**: 
   - Some methods may be more suitable for certain problem types
   - Use domain knowledge to guide method selection

4. **Document Findings**: 
   - Record which algorithms were compared
   - Note correlation values
   - Report any inconsistencies

## Tips

- Start with moderate perturbation ranges (0.2-0.3)
- Use trajectory plots to identify problematic alternatives
- Focus on criteria with high sensitivity
- Compare multiple algorithms to validate results
- Document your findings for reproducibility

## Summary

| Class | Purpose | Key Methods |
|-------|---------|-------------|
| `WeightPerturbation` | Sensitivity analysis | `weight_perturb()`, `perturb_rank()`, `perturb_stab_plot()`, `perturb_corr_heatmap()` |
| `AlgoRankCorrelation` | Algorithm comparison | `calculate()`, `print()` |


